SunPy - Python for Solar Physics

### 2014-01-26 3-5-Database.tex
rewrite of the database section

\subsection{The SunPy Database}
% TODO: reference / footnote to Summer of Code website?
The database package has been added as a Google Summer of Code (GSoC) project
in 2013. Its purpose is to offer a database interface to store and manage data
that has been retrieved via packages from the net package. Note that currently
only the VSO package is supported, not the HEK package.

The need for a database package arose because using VSO alone led to two major
problems:
\begin{enumerate}
  \item downloaded data cannot be queried or browsed
  \item to fetch data that has already been saved on the local hard drive,
    the VSO has to be queried again
\end{enumerate}

Here is a list of the main features:
\begin{itemize}
  \item detects if certain query has already been used before and queries
    the database then instead of the VSO
  % TODO: reference to SA docs? or list all currently supported SQL dialects here?
  \item many supported SQL dialects
  % TODO: footnotes for LRU/LFU or some explanation in parens behind them?
  \item database may act as a Cache (algorithm may be LRU, LFU or a custom one)
  \item all writing operations can be undone and redone
\end{itemize}

Using the database package, one uses the Database class to connect with a
database and perform operations on it.

\begin{listing}
\begin{minted}{python}
>>> from sunpy.net import vso
>>> from sunpy.database import Database
>>> database = Database('sqlite:///')
>>> database.download(
...     vso.attrs.Time((2011, 9, 20, 1), (2011, 9, 20, 2)),
...     vso.attrs.Instrument('RHESSI'))
>>> len(database)
6
\end{minted}
\caption{Connecting to a database and adding new entries.}
\label{code:db_1}
\end{listing}

In the code listing \ref{code:db_1}, a connection to new SQLite in-memory
database is established. After that, the attributes from the \textsc{VSO}
package are used to query the \textsc{VSO} and download the corresponding
files. The function \texttt{len} can be used to get the number of all database
records.

Note that the number of new database entries does not depend on the
number of files being downloaded but on the total number of \textsc{FITS}
headers in the files! If the used query had been used to query the \textsc{VSO}
directly, the number of query results would have been 2. You can see in the
following code example, listing \ref{code:db_2}, that indeed the entries can
be grouped to two results.

The function \texttt{display\_entries} displays an iterable of database entries
in a nicely-formatted \textsc{ASCII} table. The headlines correspond to the
attributes of the respective database entries.

\begin{listing}
\begin{minted}{python}
>>> from sunpy.database.tables import display_entries
>>> print display_entries(
...     database,
...     ['id', 'observation_time_start', 'wavemin', 'wavemax', 'download_time'])
id observation_time_start wavemin        wavemax           download_time      
-- ---------------------- -------        -------           -------------      
1  2011-09-20 01:09:20    0.413280643067 7.29318781883e-05 2014-01-18 19:21:35
2  2011-09-20 01:09:20    0.413280643067 7.29318781883e-05 2014-01-18 19:21:35
3  2011-09-20 01:09:20    0.413280643067 7.29318781883e-05 2014-01-18 19:21:35
4  2011-09-19 23:33:40    0.413280643067 7.29318781883e-05 2014-01-18 19:24:53
5  2011-09-19 23:33:40    0.413280643067 7.29318781883e-05 2014-01-18 19:24:53
6  2011-09-19 23:33:40    0.413280643067 7.29318781883e-05 2014-01-18 19:24:53
\end{minted}
\caption{Displaying database entries in a table.}
\label{code:db_2}
\end{listing}

As said before, the database can be used as a cache. By default, the size of
the cache is not limited which in other words means that caching is disabled.
To limit the number of database entries to a certain number a posteriori, use
the method \texttt{set\_cache\_size} and pass it the number of maximum entries.
The default algorithm to remove entries from the cache is \textsc{LRU}
(least recently used), i.e. the item whose usage is farthest away is removed
first.  % FIXME is 'whose' correct here? ^

>>>>

\subsection{The File Database}
Easy access to large quantities of solar data frequently leads to datafiles accumulating
in local storage such as laptops and desktop computers. Keeping data organized and available
is typically a cumbersome task for the average user. The file database is a sub-package of 
Sunpy which is meant to solve this problem by providing a unified database to store and manage information about 
local data files. The database package was developed as part of the 2013 
Google Summer of Code.

The database sub-package makes use of a SQLite database supported by  
SQLAlchemy\footnote{\url{http://www.sqlalchemy.org}}. This library was chosen
since it supports many SQL dialects. The database is stored as a single file which is 
created automatically. This solution makes it possibly to easily backup, copy and distribute
a database file. A server-based database, on the other hand, could be used by a collaboration 
who work together from different machines on the
same data: a central database server stores all data and the clients connect to
it to read or write data.

The database can store and manage all data that is retrieved via sunpy from the net
package though currently only the \textsc{VSO} package is supported. 
It also possible to add entries manually by file or directory. The package also provides
a \texttt{fetch} method which provides a unified data search which includes both local files
and files on the \textsc{VSO} are searched. This reduces the likelihood of downloading the same file 
multiple times. When a file is added to the database the file is scanned and a file hash is produced. 
The current date is associated with the entry along with metadata summaries such 
as instrument, date of observation, field of view, etc. 
The database also provides the ability to associate custom meta-data to 
each database entry such as keywords, comments, and favoriting.

A very special feature of the database package is the support of \texttt{undo}
and \texttt{redo} operations. This is particularly convenient in
interactive sessions to easily revert accidental operations. 
This feature will also be desirable when creating a GUI frontend for this package.

Using the database package, one uses the \texttt{Database} class to connect
with a database and perform operations on it. Figure~\ref{code:database} shows how to connect
to an in-memory database and download data from the \textsc{VSO}. These entried are
automatically added to the database. The function \texttt{len} is used to get the number of
records. The function \texttt{display\_entries} displays an iterable of database entries
in a nicely-formatted \textsc{ASCII} table. The headlines correspond to the
attributes of the respective database entries.

\begin{listing}[H]
\begin{minted}{python}
>>> from sunpy.net import vso
>>> from sunpy.database import Database
>>> database = Database('sqlite:///')
>>> database.download(
...     vso.attrs.Time((2011, 9, 20, 1), (2011, 9, 20, 2)),
...     vso.attrs.Instrument('RHESSI'))
>>> len(database)
6
>>> from sunpy.database.tables import display_entries
>>> print display_entries(
...     database,
...     ['id', 'observation_time_start', 'wavemin', 'wavemax', 'download_time'])
id observation_time_start wavemin        wavemax           download_time      
-- ---------------------- -------        -------           -------------      
1  2011-09-20 01:09:20    0.413280643067 7.29318781883e-05 2014-01-18 19:21:35

\end{minted}
\caption{Example usage of the database sub-package.}
\label{code:database}
\end{listing}
